Accurate binding of calcium to phospholipid bilayers by effective inclusion of electronic polarization

### 2017-05-12 Manuscript/proto-manuscript.tex
Proto-MS edits after a very brief discussion with Pavel

\begin{abstract}
% insert abstract here
Classical molecular dynamics simulations give detailed information about membrane structure and dynamics. However, there is still a room for improvements in current force fields – it is known from the literature, that the binding of ions, especially cations, to phopholipid membranes is overestimated in all classical models [1]. We suggest that the membrane-ion interactions can be corrected by including implicit electronic polarizability into the lipid models through the electronic continuum correction (ECC) [2], which was already applied to monovalent and divalent ions yielding models that feature correct ion pairing [3]. Using the electrometer concept [3, 4], our simulations point out that our hypothesis is correct and ECC is indeed a missing important contribution in current classical lipid models. Moreover, the solid physical principles behind ECC are found not to hamper other relevant properties of a phospholipid bilayer. 
The new lipid model, "ECClipids", shows accurate affinity to sodium and calcium cations. 
This work will continue as an open collaboration project NMRlipids IV (http://nmrlipids.blogspot.fi).

@@

 Current force fields -- pros and cons, at a good shape in many aspects, agree on various properties.

 lipid force fields fail in description of membrane-cation interaction -- could be answered by ECC? 
Cations were shown to generally overbind in PC lipid bilayers in
NMRlipids II project. Here we propose that the cation overbinding can
be corrected by implicitly inducing electronic polarizability in
lipid headgroups by scaling the partial charges.

@@

ECC and solvation $\Delta G$.(from \cite{Leontyev2015}: hydration $\Delta G$ can provide good structure and energetics, but will fail in good interactions -- already proved for cations)
In addition, $\Delta G_{hyd}$ -- the usual target for parametrization of small molecules in classical force fields -- is not the right target in the MDEC/ECC paradigm -- part of $\Delta G$ is already included in the polarization of electrons, and only the remining part of $\Delta G$ belongs to the polarization of the nuclei. 
Hence it is non trivial, how large should be the scaling factor $f_\sigma$ -- it should lay between $1$ and $f_q$, the limits for the original interaction energy and its limit when we neglect the $\Delta G_{el}$ term in $\Delta G_{hyd}$.

How do I apply ECC on Lipid14 POPC and why such choice (i.e. $f_\sigma$, $f_q$ and definition of the scaled region).

why did I choose Lipid 14 -- good ratio of $\alpha / \beta$ response.

@@

Accurate electrometer response to \ce{Na^+}, \ce{Ca^{2+}} and \ce{DHMDMAB} cationic surfactant. 
The response of the headgroup order parameters is improved both through diminished affinity towards the cations and through lower headgroup response to the bound charge (from simulations with \ce{DHMDMAB} surfactant). 
The best models from Fig.  \ref{OrderParameterIONSchangesSCALED} are shown in
Fig. \ref{OrderParameterCHANGESnewMODELS}. Based on these results we choose

Preserves other relevant properties: APL, Order Parameters (OPs), diffusion? \todoi{diffusion can be even improved}, elasticity? \todoi{check elastic properties -- do they change, improve, worsen??} -- however, it is still garbage in-garbage out
The glycerol backbone and headgroup order parameters are plotted in
Fig \ref{HGopsNEWmodels.eps} for reference models and EEC models. The difference,
i.e. the effect of scaling on these parameters, is shown in Fig. \ref{HGopsNEWmodelsCHANGE}.

Gives correct scattering form factor -- good ground for interpretting experimental data.

Provides/predicts correct stoichiometry for \ce{Na^+}, \ce{Ca^{2+}} and their interaction energies with the lipid membrane.

@@

other lipids: charged lipid? -- ongoing research in our lab (starting to do something with Aniket on POPE for curved membranes)

@@

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Fig/OrderParameterIONSchangesSCALED.eps}
  \caption{\label{OrderParameterIONSchangesSCALED}
    Headgroup order parameter changes as a function of cations from models
    with modified headgroup partial charges.}
\end{figure}

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Fig/HGopsNEWmodelsCHANGE.eps}
  \caption{\label{HGopsNEWmodelsCHANGE}
    Changes in headgroup and glycerol backbone order parameters due to EEC.}
\end{figure}

\begin{table}
  \caption{Area per lipid from different models for POPC without ions\label{apls} }
  \begin{tabular}{c c}
    model          & A (Å$^2$)    \\
    Lipid14 (literature)  & 65.6$\pm$ 0.5   \\
    Lipid14ecc0.80+sigma0.875 & 64.3      \\
    Slipids (literature T=303K)       & 64.6 $\pm$ 0.4   \\
    SlipidsEEC0.8  & 57.8  \\
  \end{tabular}
\end{table}

>>>>

\begin{abstract}
% insert abstract here
Classical molecular dynamics simulations give detailed information about membrane structure and dynamics. 
However, there is still a room for improvements in current force fields – it is known from the literature, that the binding of ions, especially cations, to phopholipid membranes is overestimated in all classical models [1]. 
We suggest that the membrane-ion interactions can be corrected by including implicit electronic polarizability into the lipid models through the electronic continuum correction (ECC) [2], which was already applied to monovalent and divalent ions yielding models that feature correct ion pairing [3]. 
Using the electrometer concept [3, 4] and x-ray scattering form factors, our simulations point out that our hypothesis is correct and ECC is indeed a missing important contribution in current classical lipid models. 
Moreover, the solid physical principles behind ECC are found not to hamper other relevant properties of a phospholipid bilayer. 
The new lipid model, "ECClipids", shows accurate binding affinity to sodium and calcium cations and headgroup order parameter response to bound charge. 
We also provide for the first time a realistic stochiometry of bound calcium cations to a POPC membrane, and their binding sites. 
This work will continue as an open collaboration project NMRlipids IV (http://nmrlipids.blogspot.fi).

@@

 Current force fields -- pros and cons, at a good shape in many aspects, agree on various properties.  -- write basic ideas from the lipid-FF whitepaper I did recently.

 lipid force fields fail in description of membrane-cation interaction -- could be answered by ECC? 
Cations were shown to generally overbind in PC lipid bilayers in NMRlipids II project. 
Here we propose that the cation overbinding can be corrected by implicitly inducing electronic polarizability in lipid headgroups by scaling the partial charges -- i.e. MDEC/ECC~\cite{Leontyev2015}\todo{Bulid BibTex references database.} paradigm.

@@

ECC and solvation $\Delta G$.(from \cite{Leontyev2015} \todoi{Pavel: do not replicate Stuchebrukhov paper, write it very briefly}: hydration $\Delta  G_{hyd}$ can provide good structure and energetics, but will fail in good interactions -- already proved for cations)
In addition, $\Delta G_{hyd}$ -- the usual target for parametrization of small molecules in classical force fields -- is not the right target in the MDEC/ECC paradigm -- part of $\Delta G = \Delta G_{nuc} + \Delta G_{el}$ is already included in the polarization of electrons, and only the remining part of $\Delta G$,  $\Delta G_{nuc}$, belongs to the polarization of the nuclei. 
Hence it is non trivial, how large should be the scaling factor $f_\sigma$ -- it should lay between $1$ and $f_q$, the limits for the original interaction energy and its limit when we neglect the $\Delta G_{el}$ term in $\Delta G_{hyd}$.

How do I apply ECC on Lipid14 POPC and why such choice (i.e. $f_\sigma$, $f_q$ and definition of the scaled region).
Region definition: headgroup requires optimization, tails appear already accurate enough. 
Choice of $f_q$: value 0.8 reflects the fact that the charges in Lipid14 were derived in vacuum, whereas they should rather be the average of vacuum and solvated charges (so called IPolQ charges \cite{Cerutti2013}, both charge-sets can be taken from \cite{Maciejewski2014a}).

why did I choose Lipid 14 -- good ratio of $\alpha / \beta$ response~\cite{NMRlipid2paper}.

@@

Accurate binding of \ce{Na^+} and \ce{Ca^{2+}} cations.

Accurate electrometer response to \ce{Na^+}, \ce{Ca^{2+}} and \ce{DHMDMAB} cationic surfactant.

What affects the headgroup OP response?: The response of the headgroup order parameters is improved both through diminished affinity towards the cations and through lower headgroup response to the bound charge (from simulations with \ce{DHMDMAB} surfactant). 
The responses of headgroup OPs are shown in Fig. \ref{OrderParameterCHANGESnewMODELS}.

Gives correct scattering form factor -- reliable structural properties (especially in combination with OPs) and good ground for interpretting experimental data.

Provides/predicts correct stoichiometry for \ce{Na^+}, \ce{Ca^{2+}} and their interaction energies with the lipid membrane.
Samuli: the main selling argument would be that we solve the details (stoichiometry, coordination numbers, etc.) of cation binding in PC bilayer.

Preserves other relevant properties: APL, Order Parameters (OPs of headgroup and tails), 
diffusion? \todoi{diffusion can be even improved -- and doesn't have to. We may as well omit it.}, 
elasticity? \todoi{check elastic properties -- do they change, improve, worsen?? -- likely not crucial for the main message of the paper.} 
-- however, it is still garbage in-garbage out

The glycerol backbone and headgroup order parameters are plotted in
Fig \ref{HGopsNEWmodels.eps} for reference models and EEC models. The difference,
i.e. the effect of scaling on these parameters, is shown in Fig. \ref{HGopsNEWmodelsCHANGE}.

\begin{table}
  \caption{Area per lipid from different models for POPC without ions\label{apls} }
  \begin{tabular}{c c c}
    model          & A (Å$^2$)   & Temperature [K] \\
    Lipid14 (literature)  & 65.6$\pm$ 0.5  &  303 \\
    Lipid14ecc0.80+sigma0.875 & 63.7 -- 64.9 &  313    \\
    experiment \cite{Jambeck2012}\todoii{REF}{put original references, not Slipids param. paper.}  & 62.7  &  293    \\
    experiment  & 64.3  &  303    \\
    experiment  & 67.3  &  323    \\
    experiment  & 68.1  &  333    \\
    experiment POPE  & 56.6 &  303    \\
  \end{tabular}
\end{table}

\textbf{Required figures}

- Figure showing the system: POPC, membrane, atoms involved in headgroup OPs.\todoi{This is probably sufficient as the last thing we would do -- and will probably provide a good basis for TOC graphic.}

- Figure comparing order parameters (Fig.~\ref{HGopsNEWmodels}) and form factor \todoii{Add plot}{Add the form factor plot} between original Lipid14, new model and experiments,

- Figure showing order parameters as a function of \ce{NaCl} and \ce{CaCl_2} concentrations~
\todoii{ongoing}{Actual concentration of cations in simulation has yet to be estimated. 
If it varies too much from the nominal concentration, I may need to tweak the scaling factors, $f_q$ or only $f_\sigma$, to accomodate it. 
However, it is very unlikely, response to the surfactant \ce{DHMDMAB} is OK. 
Big patches with loads of solvent are running at the moment to guide me on the possible finite-size errors and this conc-error.} 
from Lipid14, new model and experiments (like current Fig. 1, but only original and final models)

- Figure(s) analysing binding details (density profiles, lipids bound per ion, something else?).

@@

other lipids: charged lipids? -- ongoing research in our lab.
I'm currently working on POPE with Aniket for curved (and flat) membranes.

Samuli: From the point of view of this paper, the most relevant other lipid to study would be DPPC to see if we can reproduce the difference between POPC and DPPC in experiments.

Joe: In addition, there are more experimental OP data on DPPC. However, it is not necessary and we can leave it to the community project.

@@
